---
title: "Math Camp"
subtitle: "Northwestern University -- Social Science -- Summer 2026"
authors:
  - "[Artur Baranov](https://artur-baranov.github.io)"
  - "[Gustavo Diaz](https://gustavodiaz.org/)"
  - "TBD"
format: 
  html:
    page-layout: full
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(tinytable)
library(markdown)
```

## Welcome

<!-- Duke math camp syllabus  -->
<!-- https://sites.duke.edu/daveasiegel/files/2020/07/Mathematics-for-Political-and-Social-Research-Syllabus_Siegel-1.pdf -->

This is the website for the 2025 Social Science Math Camp, hosted by the Departments of [Political Science](https://polisci.northwestern.edu/) and [Sociology](https://sociology.northwestern.edu/) at Northwestern University. The purpose of Math Camp is to give you a chance to brush up on concepts and skills that underlie the practice of quantitative data analysis in the social sciences.

Our schedule features math in the morning and statistical programming in the afternoon. During lunch, we will also have a chance to meet faculty and graduate students and have more informal conversations about the role of methods training in graduate school.

<!-- Here are the slides from the information session on [Research Computing and Data Services resources](rcds.pptx). -->

### Details

- {{< fa calendar >}} &nbsp; September 9 -- 18
- {{< fa clock >}} &nbsp; 9:30 AM -- 4:00 PM
- {{< fa location-dot >}} &nbsp; TBD

## Schedule


```{r schedule, include = FALSE}
schedule =
  tribble(
    ~Day, ~Date, ~Morning, ~Afternoon,
    1, "Wednesday, September 9", "Introduction", "R and RStudio",
    2, "Thursday, September 10", "Notation, sets, functions", "Tidyverse I",
    3, "Friday, September 11", "Matrices", "Tidyverse II",
    4, "Monday, September 14", "Calculus I", "Sampling and simulation",
    5, "Tuesday, September 15", "Calculus II", "Markdown",
    6, "Wednesday, September 16", "Probability", "NO MEETING",
    7, "Thursday, September 17", "GitHub", "NO MEETING ",
    8, "Friday, September 18", "Review", "NO MEETING"
)
```


```{r show-schedule, echo = FALSE, results = "asis"}
schedule |>  
  tt() |> 
  format_tt(markdown = TRUE)
```


<!-- For my own reference, the order of lunch guests: -->
<!-- Wed Sept 3: Tanner, Felix-->
<!-- Thu Sept 4: Alex, Doron-->
<!-- Fri Sept 5: Isa, Liz-->
<!-- Tu Sept 9: Emilio Lehoucq -->
<!-- Fri Sept 12: Andres -->




