---
title: "Matrices"
author: "NU Social Science Math Camp"
format: 
  revealjs:
    embed-resources: true
    slide-number: false
    progress: false
---

## Getting ready

```{r setup, include = FALSE}
library(tidyverse)
library(tinytable)

knitr::opts_chunk$set(fig.pos = "center", echo = TRUE, 
               message = FALSE, warning = FALSE)
```

- Set up [eduroam](https://services.northwestern.edu/TDClient/30/Portal/KB/ArticleDet?ID=1851)

- Go back and make sure you have the MOST RECENT version of R and RStudio

## Agenda

- **Morning:** Matrices and matrix algebra
- **Lunch:** Senior faculty
- **Afternoon:** Tidyverse I

# Matrices

## Scalars

A single number is a scalar

```{r}
a = 12
a
```

$$a = 12$$

## Vectors

We already talked about vectors in R. 

```{r}
b = c(12,14,15)
```

How would you define them?

. . .

A collection of *concatenated* elements

## Vectors

Column vector

$$
\overrightarrow b = 
\begin{bmatrix}
  12 \\
  14 \\
  15
\end{bmatrix}
$$

. . .

Row vector

$$
\overrightarrow c = \begin{bmatrix}
  12 & 14 & 15
\end{bmatrix}
$$

## Vectors in R

```{r}
c(5, 25, -2, 1)
```

&nbsp;

. . .

```{r, eval = FALSE}
10:20
```


## Vectors in R

```{r}
c(5, 25, -2, 1)
```

&nbsp;

```{r}
10:20
```

&nbsp;

. . .

```{r}
seq(from = 3, to = 27, by = 3)
```

