---
title: "Calculus I "
author: "NU Social Science Math Camp"
format: 
  revealjs:
    slide-number: false
    progress: false
    code-line-numbers: false
    code-overflow: wrap
---

## What have we learned so far?

Foundational ideas behind what you will have to deal with during your graduate program:

::: incremental
- Reading about methods $\rightarrow$ Notation, sets, functions
- Working with *data frames* $\rightarrow$ Matrix operations
:::

```{r setup, include = FALSE}
library(tidyverse)
library(tinytable)

knitr::opts_chunk$set(fig.pos = "center", echo = FALSE, 
               message = FALSE, warning = FALSE)

# ggplot global options
theme_set(theme_gray(base_size = 20))
```

## Up next

Dealing with **continuity** (which implies *infinity*) $\rightarrow$ Calculuss

. . .

Two new operators:

1. Derivatives **(Today)**

2. Integrals **(Friday)**

## Calculus

::: incremental
- This is often where people experience the most difficulty

- We are now operating with whole functions!

- New operators have new, different rules

- We will stick together!

:::

## Derivatives as a slope

:::: {.columns}

::: {.column width="70%"}

```{r}
ggplot() +
  stat_function(fun = function(x){2 * x}, 
                xlim = c(-10, 10), 
                linewidth = 3) +
  theme_gray(base_size = 30)
```

:::

::: {.column width="30%"}

::: incremental
- How do we find the slope of this line?

- $m = \frac{f(x_2) - f(x_1)}{x_2-x_1} = \frac{rise}{run}$

- Let's try it out!

:::

:::

::::

## Derivatives as rate of change

::: incremental
- Positive derivative $\rightarrow$ **Function is increasing**

- Negative derivative $\rightarrow$ **Function is decreasing**

- Zeroes $\rightarrow$ Max or min (more soon!)
:::

## Now try this one!

$y = x^2$

```{r}
ggplot() +
  stat_function(fun = function(x){x ^ 2}, 
                xlim = c(-10, 10),
                linewidth = 3)
```


## Why doesn't it work?

::: incremental
- We need an operation to calculate the slope or *rate of change* for (almost) any kind of *continuous* function

- Enter **derivatives**
:::

. . .

Generalize slope formula

$$
m = \frac{f(x_1+ \Delta x) - f(x_1)}{\Delta x}
$$

## Derivatives

When $\Delta x$ is small enough

$$
\lim_{\Delta x\to0} \frac{f(x_1+ \Delta x) - f(x_1)}{\Delta x} = \frac{d}{dx} f(x) = f'(x) 
$$

::: incremental
- Read $\frac{d}{dx} f(x)$ as "the derivative of $f$ of $x$ with respect to $x$"

- Read $f'(x)$ as "$f$ prime $x$" for a shorthand

:::

## How to compute a derivative?

::: incremental
- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- Solve analytically
- Software
:::

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- **Use properties of limits**
- Solve analytically
- Software

## Using limits

$$
\lim_{\Delta x\to0} \frac{f(x_1+ \Delta x) - f(x_1)}{\Delta x}
$$

## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$

. . .

Back to $x^2$

. . .

$$
\lim_{h \to0} \frac{f(x+ h)^2 - x^2}{h}
$$

## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$


Back to $x^2$


$$
\lim_{h\to0} \frac{f(x+ h)^2 - x^2}{h} = \lim_{h\to0} \frac{x^2 + 2xh + 2h^2 - x^2}{h}
$$

## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$


Back to $x^2$


$$
\lim_{h\to0} \frac{f(x+ h)^2 - x^2}{h} = \lim_{h\to0} \frac{x^2 + 2xh + 2h^2 - x^2}{h} = \lim_{h \to 0} \frac{2xh + 2h^2}{h}
$$



## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$


Back to $x^2$


$$
\lim_{h\to0} \frac{f(x+ h)^2 - x^2}{h} = \lim_{h\to0} \frac{x^2 + 2xh + 2h^2 - x^2}{h} = \lim_{h \to 0} \frac{2xh + 2h^2}{h}
$$
$$
= \lim_{h \to 0} \frac{(2x+2h) h}{h}
$$

## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$


Back to $x^2$


$$
\lim_{h\to0} \frac{f(x+ h)^2 - x^2}{h} = \lim_{h\to0} \frac{x^2 + 2xh + 2h^2 - x^2}{h} = \lim_{h \to 0} \frac{2xh + 2h^2}{h}
$$
$$
= \lim_{h \to 0} \frac{(2x+2h) h}{h} = \lim_{h \to 0} 2x+2h
$$


## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$


Back to $x^2$


$$
\lim_{h\to0} \frac{f(x+ h)^2 - x^2}{h} = \lim_{h\to0} \frac{x^2 + 2xh + 2h^2 - x^2}{h} = \lim_{h \to 0} \frac{2xh + 2h^2}{h}
$$
$$
= \lim_{h \to 0} \frac{(2x+2h) h}{h} = \lim_{h \to 0} 2x+2h
$$
$$
=2x+0
$$

## Using limits

$$
\lim_{h\to0} \frac{f(x_1+ h) - f(x_1)}{h}
$$


Back to $x^2$


$$
\lim_{h\to0} \frac{f(x+ h)^2 - x^2}{h} = \lim_{h\to0} \frac{x^2 + 2xh + 2h^2 - x^2}{h} = \lim_{h \to 0} \frac{2xh + 2h^2}{h}
$$
$$
= \lim_{h \to 0} \frac{(2x+2h) h}{h} = \lim_{h \to 0} 2x+2h
$$
$$
=2x+0 = 2x
$$

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- **Use properties of limits**
- Solve analytically
- Software

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- **Solve analytically**
- Software


## Rules of differentiation

. . .

**Differentation:** "Taking the derivative"

## Rules of differentiation

**Constant rule:** $(c)' = 0$.

. . .

$y = 2$

```{r}
ggplot() +
  stat_function(fun = function(x){2}, xlim = c(-10, 10),
                linewidth = 3)
```

## Rules of differentiation

<!-- Demonstrate by hand -->

**Coefficient rule:** $(c \cdot f(x))' = c \cdot f'(x)$

. . .

```{r}
ggplot() +
  stat_function(fun = function(x){2 * x}, xlim = c(-10, 10), aes(color = "y = 2x"), linewidth = 3) +
  stat_function(fun = function(x){4 * x}, xlim = c(-10, 10), aes(color = "y = 4x"), linewidth = 3) +
  scale_color_manual("Function", values = c("red", "blue"))
```

## Rules of differentiation

**Sum/difference rule:** $(f(x) \pm g(x))' = f'(x) \pm g'(x)$.

. . .

Example: $f(x) = 3x + 7$

. . .

$$
f'(x) = (3x + 7)'
$$

. . .

$$
= (3x)' + (7)'
$$

. . .

$$
= 3 + 0
$$

. . .

$$
= 3
$$

## Rules of differentiation

**Power rule:** $(x^n)'=nx^{(n-1)}$

. . .

$$
f(x) = x^2
$$

. . .

$$
f'(x) = nx^{(n-1)}
$$

. . .

$$
= 2x^{(2-1)}
$$

. . .

$$
= 2x
$$

::: incremental
- Rate of change depends on values of $x$
- "... at a given point"
:::

## Practice

Let's try out $\frac{d}{dx}4x^3$ and $\frac{d}{dx}(x^2 + 2x)$ on the board.

## More practice

Use the differentiation rules we have covered so far to calculate the derivatives of $y$ with respect to $x$ of as many as these as you can.

1)  $y = 2x^2 + 10$
2)  $y = 5x^4 - \frac{2}{3}x^3$
3)  $y = 9 \sqrt x$
4)  $y = \frac{4}{x^2}$
5)  $y = \frac{2w}{5}$

## Rules of differentiation

**Exponent and logarithm rules:**

$$
\begin{aligned}
(c^x)' &= c^x \cdot ln(c), & \forall x>0 \\
\\
(log_a(x))' &= \frac{1}{x \cdot ln(a)}, & \forall x>0  \\
\end{aligned}
$$

## Rules of differentiation

**Exponent and logarithm rules:**

[Euler's number](https://tinyurl.com/2j4a9ykc) ($e$) has interesting properties when it comes to derivatives.

$$
\begin{aligned}
(e^x)' &= e^x \\
\\
(ln(x))' &= \frac{1}{x}, & \forall x>0 
\end{aligned}
$$

## More more practice

Compute the following:

1)  $\frac{d}{dx}(10e^x)$
<!-- 10e^x -->
2)  $\frac{d}{dx}(ln(x) - \frac{e^2}{3})$
<!-- 1/x -->

## Advanced rules

**Product rule:** $(f(x)g(x))'=f'(x)g(x) + g'(x)f(x)$

. . .

$$
[(3x+4)(x+2)]'
$$

. . .


$$
f(x) = (3x+4), g(x) = (x+2); f'(x) = 3, g'(x) = 1
$$

. . .

$$
\color{blue}{f'(x)}\color{yellow}{g(x)} + \color{purple}{g'(x)}\color{green}{f(x)}
$$

. . .

$$
= \color{blue}3 \color{yellow}{(x+2)} + \color{purple}{1} \color{green}{(3x+4)}
$$

. . .

$$
= 3x + 6 + 3x+ 4
$$

. . .

$$
= 6x + 10
$$

## Advanced rules

**Quotient rule:** $\displaystyle(\frac{f(x)}{g(x)})' = \frac{f'(x)g(x) + g'(x)f(x)}{[g(x)]^2}$

. . .

$$
\frac{3x^2}{x+2}
$$

## Advanced rules

**Quotient rule:** $\displaystyle(\frac{f(x)}{g(x)})' = \frac{f'(x)g(x) + g'(x)f(x)}{[g(x)]^2}$

$$
\frac{3x^2}{x+2} \Rightarrow f(x) = 3x^2, g(x) = x + 2; f'(x) = 6x, g'(x) = 1
$$

. . .

$$
\frac{\color{blue}{f'(x)}\color{yellow}{g(x)} + \color{purple}{g'(x)}\color{green}{f(x)}}{[\color{yellow}{g(x)}]^2} =
\frac{\color{blue}{6x}\color{yellow}{(x+2)} - \color{purple}1 \color{green}{(3x^2)}}{\color{yellow}{(x+2)}^2}
$$

. . .

$$
= \frac{6x^2 + 12x - 3x2}{(x+2)^2} = \frac{3x^2+12x}{(x+2)^2}
$$


## Advanced rules

**Chain rule:** $(f(g(x))' = f'(g(x)) \cdot g'(x)$

. . .

$$
f(g(x)) = (x+3)^3
$$

## Advanced rules

**Chain rule:** $[f(g(x))]' = f'(g(x)) \cdot g'(x)$

$$
f(g(x)) = (x+3)^3 \Rightarrow f(g(x)) = g(x)^3, g(x) = (x+3)
$$

. . .

$$
f'(x) = 3x^2 \Rightarrow f'(g(x)) = 3 (x+3)^2, g'(x) = 1
$$

. . .

Substitute

$$
[f(g(x))]' = 3(x+3)^2 \cdot 1
$$

## More more more practice

1)  $\frac{d}{dx}x^3 \cdot x$
2)  $\frac{d}{dx}e^x \cdot x^2$
3)  $\frac{d}{dx}(3x^4-8)^2$

## Higher-order derivatives

::: incremental
- $f'(x)$: First derivative
- $f''(x)$: Second derivative
- $f'''(x)$?
:::

. . .

If we were driving a car:

-   $f(x)$ = distance traveled at time $x$
-   $f'(x)$ = speed at time $x$
-   $f''(x)$ = acceleration at time $x$

. . .

To get $f''(x)$ just take the derivative of the output of $f'(x)$

## Partial derivative

. . .

What happens when we have more than one variable? $f(x,z)$

. . .

Partial derivative

$$ 
\frac{\partial}{\partial_x}f(x,z) = \frac{\partial_y}{\partial_x} = \partial_x f
$$

. . .

Just treat other variables as constants and calculate the derivative of the target variable (usually $x$)

## Example

$$
\begin{aligned}
y = f(x,z) &= xz \\
\frac{\partial_y}{\partial_x} &= z
\end{aligned}
$$

. . .

Let's try $\displaystyle\frac{\partial (x^2y+xy^2-x)}{\partial x}$ and $\displaystyle\frac{\partial (x^2y+xy^2-x)}{\partial y}$ together

## Differentiability of functions

::: incremental
- Not all functions are differentiable
- We need *continuous* and *reasonably smooth* functions
- **Informally:** Can draw them without lifting the pen
- **Formally:** A function is continuous at a point $a$ if: $\lim_{x \to a} f(x)=f(a)$
- Differentiable $\to$ continuous
- Continuous $\not\to$ differentiable
:::

---

```{r, message=FALSE, warning=FALSE}
ggplot() +
  stat_function(fun = function(x){abs(x) + 2}, xlim = c(-4, 4), 
                aes(color = "y = |x| + 2"), linewidth = 3) +
  stat_function(fun = function(x){sqrt(abs(x)) + 1}, xlim = c(-4, 4), 
                aes(color = "y = √(|x|) + 1"), linewidth = 3) +
  stat_function(fun = function(x){sign(x) * abs(x)^(1 / 3)}, xlim = c(-4, 4), 
                aes(color = "y = x^(1/3)"), linewidth = 3) +
  scale_colour_manual("Function", values = c("red", "blue", "black")) +
  labs(title = "Examples of functions that are not differentiable at x=0")
```

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- **Solve analytically**
- Software

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- Solve analytically
- **Software**

## Using computers

::: incremental
- **Symbolic differentiation:** Do the math for you ([WolframAlpha](https://www.wolframalpha.com/calculators/derivative-calculator/) or [Mathematica](https://www.wolfram.com/mathematica/))
- **Numerical differentiation:** Evaluate the function at many values (`optim()` in R)
- **Automatic differentiation:** Break up functions into elementary operations and use chain rule (`torch` package in R, various packages in Python)
:::


## Optimization

::: incremental
- Find minimum or maximum values

- Valleys are *minima*

- Peaks are *maxima*

- Both are *extrema*

- Extrama can be *local* or *global*
:::

## First order condition

::: incremental

- **Minimum:** Derivative is negative before and positive after

- **Maximum:** Derivative is positive before and negative after

- Extrema are *critical points* 

- **First order condition** helps us find them
:::

. . .

$$FOC: f'(x)=0$$

## Second order condition

. . .

FOC finds critical points, but can't tell if minimum or maximum

. . .

**Second derivative test**

::: incremental
-   Start by identifying $f''(x)$

-   Substitute in the stationary points $(x^*)$ identified from the FOC

-   $f''(x^*) > 0$ we have a local minimum

-   $f''(x^*) < 0$ we have a local maximum

-   $f''(x^*) = 0$ we (may) have an inflection point - need to calculate higher-order derivatives
:::

## More more more more practice

Obtain FOC and SOC for $\displaystyle y=\frac{1}{2} x^3 + 3 x^2 - 2$

```{=html}
<iframe src="https://www.desmos.com/calculator/qccgx2ukl7?embed" width="1000" height="500" style="border: 1px solid #ccc" frameborder=0></iframe>
```


## Local vs. global extrema

**Extreme value theorem:**

::: incremental
- If a real-valued function is *continuous on a closed and bounded (finite) interval*, the function *must have a global minimum and a global maximum* on that interval at least once

- Either at the boundaries or at one of the local extrema

- Find by comparing local min/max to the value of the function at the endpoints

- Could be limits as function approaches $-\infty$ and $+\infty$
:::

## Practice again

Check $\displaystyle y=\frac{1}{2} x^3 + 3 x^2 - 2$ for global extrema

```{=html}
<iframe src="https://www.desmos.com/calculator/qccgx2ukl7?embed" width="1000" height="500" style="border: 1px solid #ccc" frameborder=0></iframe>
```

## Another one

Identify the global extrema of the function $\displaystyle \frac{x^3}{3} - \frac{3}{2}x^2 -10x$ in the interval $[-6, 6]$

<iframe src="https://www.desmos.com/calculator/wetagmqrig?embed" width="800" height="400" style="border: 1px solid #ccc" frameborder=0></iframe>
