---
title: "Calculus I"
author: "NU Social Science Math Camp"
format: 
  revealjs:
    embed-resources: true
    slide-number: false
    progress: false
    code-line-numbers: false
    code-overflow: wrap
---

## What have we learned so far?

Foundational ideas behind what you will have to deal with during your graduate program:

::: incremental
- Reading about methods $\rightarrow$ Notation, sets, functions
- Working with *data frames* $\rightarrow$ Matrix operations
:::

```{r setup, include = FALSE}
library(tidyverse)
library(tinytable)

knitr::opts_chunk$set(fig.pos = "center", echo = FALSE, 
               message = FALSE, warning = FALSE)

# ggplot global options
theme_set(theme_gray(base_size = 20))
```

## Up next

Dealing with **continuity** (which implies *infinity*) $\rightarrow$ Calculus

. . .

Two new operators:

1. Derivatives **(Today)**

2. Integrals **(Friday)**

## Calculus

::: incremental
- This is often where people experience the most difficulty

- We are now operating with whole functions!

- New operators have new, different rules

- We will stick together!

:::

## Derivatives as a slope

:::: {.columns}

::: {.column width="70%"}

```{r}
ggplot() +
  stat_function(fun = function(x){2 * x}, 
                xlim = c(-10, 10), 
                linewidth = 3) +
  theme_gray(base_size = 30)
```

:::

::: {.column width="30%"}

::: incremental
- How do we find the slope of this line?

- $m = \frac{f(x_2) - f(x_1)}{x_2-x_1} = \frac{rise}{run}$

- Let's try it out!

:::

:::

::::

## Now try this one!

$y = x^2$

```{r}
ggplot() +
  stat_function(fun = function(x){x ^ 2}, 
                xlim = c(-10, 10),
                linewidth = 3)
```
