---
title: "Calculus II"
author: "NU Social Science Math Camp"
format: 
  revealjs:
    slide-number: false
    progress: false
    code-line-numbers: false
    code-overflow: wrap
    chalkboard: true
---

## Last time
### How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- Solve analytically
- Software

```{r setup, include = FALSE}
library(tidyverse)
library(tinytable)

knitr::opts_chunk$set(fig.pos = "center", echo = FALSE, 
               message = FALSE, warning = FALSE)

# ggplot global options
theme_set(theme_gray(base_size = 20))
```


## Last time
### How to compute a derivative?

- **Plug in many values into $\frac{rise}{run}$**
- **Use properties of limits**
- **Solve analytically**
- Software

## Today

- Wrap up derivatives
- What makes derivatives useful?
- Integrals

## Higher-order derivatives

::: incremental
- $f'(x)$: First derivative
- $f''(x)$: Second derivative
- $f'''(x)$?
:::

. . .

If we were driving a car:

-   $f(x)$ = distance traveled at time $x$
-   $f'(x)$ = speed at time $x$
-   $f''(x)$ = acceleration at time $x$

. . .

To get $f''(x)$ just take the derivative of the output of $f'(x)$

## Partial derivative

. . .

What happens when we have more than one variable? $f(x,z)$

. . .

Partial derivative

$$ 
\frac{\partial}{\partial_x}f(x,z) = \frac{\partial_y}{\partial_x} = \partial_x f
$$

. . .

Just treat other variables as constants and calculate the derivative of the target variable (usually $x$)

## Example

$$
\begin{aligned}
y = f(x,z) &= xz \\
\frac{\partial_y}{\partial_x} &= z
\end{aligned}
$$

. . .

Let's try $\displaystyle\frac{\partial (x^2y+xy^2-x)}{\partial x}$ and $\displaystyle\frac{\partial (x^2y+xy^2-x)}{\partial y}$ together

## Differentiability of functions

::: incremental
- Not all functions are differentiable
- We need *continuous* and *reasonably smooth* functions
- **Informally:** Can draw them without lifting the pen
- **Formally:** A function is continuous at a point $a$ if: $\lim_{x \to a} f(x)=f(a)$
- Differentiable $\to$ continuous
- Continuous $\not\to$ differentiable
:::

---

```{r, message=FALSE, warning=FALSE}
ggplot() +
  stat_function(fun = function(x){abs(x) + 2}, xlim = c(-4, 4), 
                aes(color = "y = |x| + 2"), linewidth = 3) +
  stat_function(fun = function(x){sqrt(abs(x)) + 1}, xlim = c(-4, 4), 
                aes(color = "y = √(|x|) + 1"), linewidth = 3) +
  stat_function(fun = function(x){sign(x) * abs(x)^(1 / 3)}, xlim = c(-4, 4), 
                aes(color = "y = x^(1/3)"), linewidth = 3) +
  scale_colour_manual("Function", values = c("red", "blue", "black")) +
  labs(title = "Examples of functions that are not differentiable at x=0")
```

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- Solve analytically
- Software

## How to compute a derivative?

- Plug in many values into $\frac{rise}{run}$
- Use properties of limits
- Solve analytically
- **Software**

## Using computers

::: incremental
- **Symbolic differentiation:** Do the math for you ([WolframAlpha](https://www.wolframalpha.com/calculators/derivative-calculator/) or [Mathematica](https://www.wolfram.com/mathematica/))
- **Numerical differentiation:** Evaluate the function at many values (`optim()` in R)
- **Automatic differentiation:** Break up functions into elementary operations and use chain rule (`torch` package in R, various packages in Python)
:::


## Optimization

::: incremental
- Find minimum or maximum values

- Valleys are *minima*

- Peaks are *maxima*

- Both are *extrema*

- Extrama can be *local* or *global*
:::

## First order condition

::: incremental

- **Minimum:** Derivative is negative before and positive after

- **Maximum:** Derivative is positive before and negative after

- Extrema are *critical points* 

- **First order condition** helps us find them
:::

. . .

$$FOC: f'(x)=0$$

## Second order condition

. . .

FOC finds critical points, but can't tell if minimum or maximum

. . .

**Second derivative test**

::: incremental
-   Start by identifying $f''(x)$

-   Substitute in the stationary points $(x^*)$ identified from the FOC

-   $f''(x^*) > 0$ we have a local minimum

-   $f''(x^*) < 0$ we have a local maximum

-   $f''(x^*) = 0$ we (may) have an inflection point - need to calculate higher-order derivatives
:::

## Practice

Obtain FOC and SOC for $\displaystyle y=\frac{1}{2} x^3 + 3 x^2 - 2$

```{=html}
<iframe src="https://www.desmos.com/calculator/qccgx2ukl7?embed" width="1000" height="500" style="border: 1px solid #ccc" frameborder=0></iframe>
```


## Local vs. global extrema

**Extreme value theorem:**

::: incremental
- If a real-valued function is *continuous on a closed and bounded (finite) interval*, the function *must have a global minimum and a global maximum* on that interval at least once

- Either at the boundaries or at one of the local extrema

- Find by comparing local min/max to the value of the function at the endpoints

- Could be limits as function approaches $-\infty$ and $+\infty$
:::

## Practice again

Identify the global extrema of the function $\displaystyle \frac{x^3}{3} - \frac{3}{2}x^2 -10x$ in the interval $[-6, 6]$

<iframe src="https://www.desmos.com/calculator/wetagmqrig?embed" width="800" height="400" style="border: 1px solid #ccc" frameborder=0></iframe>

# Integrals

## Integration

- Inverse operation to the **derivative**

- **Derivative:** instantaneous change at a given point

- **Integrals:** *Net effect* of change of a function at a range

## Area under a curve

. . .

Calculate the area under

$$
f(x) = 
\begin{cases}
\frac{1}{3} & \text{for } x \in [0, 3] \\
0 & \text{otherwise}
\end{cases}
$$

## A more complicated example

$y=x^2$

. . .

```{r}
ggplot() +
  # draw main function
  stat_function(fun = function(x){x ^ 2}, xlim = c(-2, 2),
                linewidth = 2) +
  # fill area under the curve between x = 0 and x = 1
  geom_area(mapping = aes(x = 0), stat = "function",
            fun = function(x){x ^ 2}, xlim = c(0, 1), 
            fill = "purple", alpha = 0)
```

## A more complicated example

$y=x^2$

```{r}
ggplot() +
  # draw main function
  stat_function(fun = function(x){x ^ 2}, xlim = c(-2, 2),
                linewidth = 2) +
  # fill area under the curve between x = 0 and x = 1
  geom_area(mapping = aes(x = 0), stat = "function",
            fun = function(x){x ^ 2}, xlim = c(0, 1), 
            fill = "purple")
```

## Two kinds of integrals

- Definite integral
- Indefinite integral

## Indefinite integral

$$F(x)= \displaystyle\int f(x) \text{ } dx$$

::: incremental
- $\int$ is the anti-derivative with respect to $dx$
- $F(x)$ is the inverse of $f'(x)$
:::

. . .

$$\displaystyle\frac{dF(x)}{dx} = f(x)$$

. . .

$$F' = f$$

## Toy example

$$f(x) = 1$$

. . .

$$F'(x) = 1$$

. . .

$$F(x) = x$$

. . .

$$F(x) = x + 10$$

. . .

$$F(x) = x + 100$$

## Toy example

```{r}
ggplot() +
  # draw main function
  geom_hline(yintercept = 0, linetype = "dashed") +
  stat_function(fun = function(x){x}, xlim = c(-200, 200),
                linewidth = 2, 
                aes(color = "y = x")) +
  stat_function(fun = function(x){x + 10 }, 
                xlim = c(-200, 200),
                linewidth = 2, 
                aes(color = "y = x + 10")) +
  stat_function(fun = function(x){x + 100 }, 
                xlim = c(-200, 200),
                linewidth = 2, 
                aes(color = "y = x + 100")) +
  scale_color_manual("Function", 
                     values = c("purple", "blue", "green"))
```

## Generalizing

::: incremental
- For $f(x) = F'(x) = 1$ we can write $F(x) = x + C$

- $C$ is the **constant of integration**

- Finding the inverse of a derivative is called **anti-differentiation**

:::

## Definite integral

**Indefinite integral:**

$$\displaystyle\int f(x)dx$$

. . .

**Definite integral:**

$$\displaystyle\int_{a}^{b} f(x)dx$$

## Calculating definite integrals

. . .

[The first]{style="color: gray; font-size: 15px"} **FUNDAMENTAL THEOREM OF CALCULUS**

$$\displaystyle\int_{a}^{b} f(x) \text{ } dx = F(b)-F(a) = F(x)\bigg|_{a}^{b}$$

::: incremental
- Determine $F(x)$
- Substitute in $F(b)$ and $F(a)$
- Calculate difference
:::

## Rules of integration

**Coefficient rule:** $\displaystyle \color{gray}\int c f(x)\, \color{gray}{dx} \color{gray}= c \color{gray}\int f(x)\,\color{gray}{dx}$

. . .

**Sum/difference rule:** $\displaystyle\color{gray}\int (f(x) \pm g(x))\,\color{gray}{dx} \color{gray}= \color{gray}\int f(x)\,\color{gray}{dx} \pm \color{gray}\int g(x)\,\color{gray}{dx}$

. . .

**Constant rule:** $\displaystyle\color{gray}\int c\,\color{gray}{dx} \color{gray}= cx \color{gray}{+ C}$

. . .

**Power rule:** $\displaystyle\color{gray}\int x^n\,\color{gray}{dx} \color{gray}= \frac{x^{n+1}}{n+1} \color{gray}{+ C} \qquad \forall n \neq -1$

. . .

**Reciprocal rule:**$\displaystyle\color{gray}\int \frac{1}{x}\,\color{gray}{dx} \color{gray}= \ln(x) \color{gray}{+ C}$

## Rules of integration

**Exponent and logarithm rules:**

$$
\begin{aligned}
\displaystyle \color{gray}\int e^x \,\color{gray}{dx} &\color{gray}= e^x\color{gray}{+C} \\
\displaystyle \color{gray}\int c^x \,\color{gray}{dx} &\color{gray}= \frac{c^x}{ln(c)}\color{gray}{+C}
\end{aligned}
$$

. . .

$$
\begin{aligned}
\displaystyle \color{gray}\int ln(x) \,\color{gray}{dx} &\color{gray}= x \cdot ln(x) - x\color{gray}{+C} \\
\displaystyle \color{gray}\int log_c(x) \,\color{gray}{dx} &\color{gray}= \frac{x \cdot log_c(x) - x}{log_c(x)}\color{gray}{+C} 
\end{aligned}
$$

## Integration by parts

::: aside
Analog to the product rule.
:::

$$
\displaystyle \color{gray}\int f(x)g'(x)\,\color{gray}{dx} = f(x)g(x) - \color{gray}\int f'(x)g(x)\,\color{gray}{dx}
$$

. . .

$$
\int u dv = uv - \int vdu
$$

## Integration by parts

::: aside
Analog to the product rule.
:::

$$
\displaystyle \color{gray}\int f(x)g'(x)\,\color{gray}{dx} = f(x)g(x) - \color{gray}\int f'(x)g(x)\,\color{gray}{dx}
$$


$$
\int u dv = \underbrace{uv - \int vdu}_\text{"ultraviolet voodoo"}
$$


## Example

$$ \int x e^x dx$$

. . . 

$$ \int udv = uv - \int vdu $$

. . .

$$
\color{purple}{u = x}; \text{ } \color{green}{du = 1dx}; \text{ } \color{yellow}{dv = e^xdx} ;  \text{ } \color{blue}{v = e^x}
$$

. . .

$$
\int \color{purple}x \color{yellow}{e^xdx}= \color{purple}x \color{blue}{e^x} - \int \color{blue}{e^x} \color{green}{(1dx)}
$$

. . .

$$
= xe^x - e^x + C
$$

## Integration by (U-)substitution

::: incremental
1. Have $\displaystyle \int f(g(x))g'(x)\,dx$
2. Set $u=g(x)$
3. Compute $\int f(u)\,du$
4. Replace $u$ for $g(x)$
:::

::: aside
Analog to the chain rule.
:::

## Example

$$
\int 4x (x^2 + 5)^3 dx
$$

. . .

$$
\color{purple}{u = x2 + 5}
$$

. . .

$$
\text{} \color{green}{du = 2x dx}
$$

## Example

$$
\int 4x (x^2 + 5)^3 dx
$$

$$
\color{purple}{u = x2 + 5}
$$

$$
\text{} \color{green}{du = 2x dx \Rightarrow \frac{du}{2x} = \frac{2xdx}{2x}}  
$$

## Example

$$
\int 4x (x^2 + 5)^3 dx
$$

$$
\color{purple}{u = x2 + 5}
$$

$$
\text{} \color{green}{du = 2x dx \Rightarrow \frac{du}{2x} = \frac{2xdx}{2x} \Rightarrow \frac{du}{2x} = dx}  
$$

## Replacing

$$
\int 4x \color{purple}{(x^2 + 5)}^3 \color{green}{dx}
$$

. . .

$$
\int 4x \color{purple}u^3 \color{green}{\frac{du}{2x}}
$$

. . .



$$
= 2 \int \color{purple}{u}^3 \color{green}{du}
$$

. . .

$$
= 2 \cdot \frac{\color{purple}u}{4}^4 + C
$$



## Clean up and bring back $u$

$$
= 2 \cdot \frac{\color{purple}u}{4}^4 + C
$$

. . .

$$
= \frac{1}{2} \color{purple}u^4 + C
$$

. . .

$$
\frac{1}{2} \color{purple}{(x2 + 5)}^4 + C
$$



## Don't worry!

We are not going to practice hard rules by hand

## Practice

Solve the following definite integrals

1.  $\displaystyle\int_{1}^{7} x^2 \text{ } dx$

2.  $\displaystyle\int_{1}^{10} 3x^2 \text{ } dx$

3.  $\displaystyle\int_7^7 x\text{ } dx$

4.  $\displaystyle\int_{1}^{5} 3x^2 + 2x - 7\text{ }dx$